\LARGE \bf On Physics-Based 3D Waypoint Generation for Autonomous Exploration in Aerial Robotics

### 2011-09-19 paper/On Physics-Based 3D Waypoint Generation/paper.tex
finale version des papers for review
git-svn-id: https://politbuero.dyndns.org/svn/repos/promotion@1109 f4953187-ecf7-0310-90f3-8d0fedb0aae5

\title{\LARGE \bf
On Physics-Based 3D Waypoint Generation for Autonomous Exploration in Mobile Robotics
}

@@

\author{Benjamin Adler and Jianwei Zhang\\
        University of Hamburg\\
        Fachbereich Informatik,\\
        Technische Aspekte Multimodaler Systeme (TAMS)\\
        Hamburg, Germany\\
        {\tt\small {adler,zhang}@informatik.uni-hamburg.de}%
}

@@

This paper presents a novel, physics-based approach to computing waypoints, which are subsequently being used to steer a flying platform in three-dimensional space to map an outdoor environment. Because the algorithm relies on simple physical processes, it is both easy to understand and to implement in combination with traditional data structures. Generation of efficient sensor-trajectories for maximized information gain operates directly on unorganized point-clouds, creating a perfect fit for environment mapping with commonly user LIDAR sensors or time-of-flight cameras.

We also present the algorithm by simulating its performance in a virtual outdoor scenario.

@@

Automatic model building has always been one of the most important parts of autonomous robotics. Exploration and mapping - as a subclass of this problem - has been the topic of many papers in recent years. The SLAM problem was first solved for two-dimensional mapping scenarios and later-on expanded to three-dimensional environments \cite{shen2011}.

This paper presents a novel approach on increasing the efficiency of such mapping procedures. While the implementations of SLAM have improved greatly in recent years, one aspect of mapping three-dimensional environments was given comparatively little attention: finding the next best view. Typically, SLAM algorithms have been researched and developed on mobile platforms moving on flat surfaces \cite{walthelm2011}, \cite{yamauchi1998}, \cite{joho2007}. This setup does not require a high efficiency in exploration, as it does not inherently impose time constraints. The authors of this paper, however, are implementing exploration and mapping algorithms on a flying platform with a maximum flight time of less than 15 minutes. Aiming to explore as much of the environment as possible within this limited time motivates an efficient way of finding sensor-poses (and in extension, -trajectories) that enable sensors to deliver as much information as quickly as possible. While repeatedly passing over the unknown environment in scan-line fashion results in very fast scanning, a better approach needs to be found to ensure scanning of surfaces that are shadowed from those sensor-trajectories.

This paper is organized as follows: The next section describes , investigates,

@@

Since the introduction of SLAM, much work has been done to refine and extend the algorithm for use in both two- and threedimensional environments \cite{haehnel2004}, TODO. Finding the next best view, though, has never been a part of the SLAM problem \cite{gonzalez_banos2002} and is thus rarely addressed by these works.

Generating safely reachable waypoints from previously acquired sensor data is an extension of next-best-view problem (which itself is an extension to the art-gallery problem \cite{orourke1987}), as it does not include the constraints of safe navigation and merely deals with the generation of isolated viewpoints. As sensor-poses in areas between known and unknown environment offer a good compromise between safe reachability and high information gain, frontier-based approaches are a popular method to compute new waypoints in two-dimensional space and are well documented in the literature \cite{gonzalez_banos2002}, \cite{strand2009}, \cite{yamauchi1998}, \cite{shade2011}. Unfortunately, information about exploration boundaries is hard to generate from three-dimensional pointclouds.

As our robot is holonomic, the generated waypoints and trajectories can be created with less concern of the platform's ability to catenate them due to motion constraints that other tyes of vehicles (e.g. fixed-wing UAVs) previously entailed \cite{hota2009}. The common problem of wind, appearing as both opportunity and hindrance, but mostly in the form of further constraints on feasible paths for many UAVs has spawned many contributions in the past \cite{mcneely2007}. Even though wind is still a matter of concern for our platform, it does no longer influence the design of path planning or waypoint generation, but only localization and motion control.

@@

\begin{figure}[thpb]
  \centering
  \includegraphics[width=0.5\textwidth]{images/oktokopter_photo_small}
  \caption{The experimental flying platform with mounted GNSS-antenna/receiver, processor-board, laserscanner and IMU.}
  \label{oktorendering}
\end{figure}

To save time during the construction of our platform, we used an ``Okto 2''-octocopter from the mikrokopter-project \cite{mikrokopterproject} as a base for our vehicle (see Fig. \ref{oktorendering}). Its central FlightControl (FC) processor-board is connected to eight brushless-motor controllers (BL) via an I2C bus. Employing its on-board gyros and accelerometers, the FC is programmed to stabilize the platform by itself when no other motion-control-commands are received from either the connected remote-control-receiver or the serial port. The project also developed the NaviControl (NC), an add-on hardware module that sends motion-control-commands (thrust, yaw, pitch/roll each as 8-bit signed integers) to steer the platform to waypoints. We abandoned the idea of using the NC for our purposes, as its hardware adds extra weight and its control-algorithms are neither capable of moving the platform as desired nor open-source. Instead, we fitted an Intel Atom processor-board to the platform. This board has a serial connection to the FC, which is used to send motion-control-commands and receive sensor input.

@@

The laserscanner rotates at 2400rpm and provides both falling and rising edges after each complete scan of its 270$^\circ$ field of view. This signal is connected to the GNSS receiver, which is configured to emit the GPS Time Of Week (TOW) of the event with microsecond-precision on its USB-connection to the Atom processor-board. With both poses and laserscanner-events being timestamped by the GNSS-receiver's time system, we can interpolate the vehicle's pose to the scanner's pose at the time of each scan.

@@

We initially planned to use two laserscanners, one with its scan-plane perpendicular to both the platform's groundplane and its forward-vector for scanning the environment and another scanner oriented to have its scan-plane parallel to the vehicle's ground-plane for collision-avoidance (see Fig. \ref{lidar_setup_pair}). Besides the obvious disadvantages of another 250 grams of payload and $\sim$10W of power consumption, the platform would have had to constantly pitch during flight to ensure reliable obstacle scanning and detection. Superimposing the pitching motion has proven to result in considerable amounts of work for the motors on the platform's front and back, reducing flight time. The non-constant angular pitching speed in this setup also necessitates a more precise interpolation of the vehicle's to the scanner's pose.

To avoid these drawbacks, we instead mounted a single laserscanner as depicted in Fig. \ref{lidar_setup_single}. By permanently yawing the platform in-flight, a single scanner captures both geometry for surface reconstruction and collision avoidance, provided that the world is static and the platform does not move further than the laserscanners range within one rotation. Given a constant angular yawing-speed, the disadvantage of precise interpolation of the platform's pose becomes less problematic.
% TODO: speed < lidar-range * drehgeschwindigkeit ?

@@

In order to capture sensor data representing arbitrarily complex geometry and detail, we decided to implement our algorithm on an octree-based \cite{meagher1982} data-structure. Octrees are well suited for storing information in non-uniform resolution and the process of storing and retrieving data can easily be executed in parallel, allowing for optimization of parts of our algorithms. Besides its position, each point in the octree also stores its brightness (i.e. intensity as measured by the laserscanner) and a vector pointing from that point back to the sensor. As the platform's measured orientation is less precise that its measured position, this vector's length will be used lateron to assign points recorded from further distance (and thus suffering a greater impact from platform-orientation errors) a lesser weight in the following surface reconstruction.

@@

\begin{figure*}[ht]
  \centering
    \includegraphics[width=0.4\textwidth]{images/process_side/1}\hfill
    \includegraphics[width=0.4\textwidth]{images/process_side/2}\vspace{3mm}
    \includegraphics[width=0.4\textwidth]{images/process_side/3}\hfill
    \includegraphics[width=0.4\textwidth]{images/process_side/4}
    \caption{Waypoint generation, orthogonal projection; the bounding volume is blue, detection volume is red, vehicle position is black, pointcloud for geometry reconstruction is grey and pointcloud for waypoint generation is red. Second figure shows sampling geometry interacting with pointcloud. Geometry hitting the red volume will be converted to a waypoint at the position it last hit a point from the pointcloud. The last figure shows generated waypoints in red for an initial pointcloud population.}
    \label{explanation_of_process}
\end{figure*}

Three-dimensional environment mapping is often implemented using laserscanners and time-of-flight cameras, so unorganized point clouds are a very common type of sensor data. Any sensor generating spatial occupancy information can be used with our algorithm, as the pointcloud is its only sensory input. Finding the next best view in such data can be very hard, as it does not supply any information about geometric structures such as corners, edges, surfaces and normals. In contrast to many other algorithms (\cite{joho2007}, \cite{walthelm2011}), ours does not require such information.

Prior to the mapping process, the user creates a bounding volume as a representation of the region-of-interest around the vehicle's starting position, thereby defining the environment that is to be mapped. All generated viewpoints will be constrained to this volume and its vicinity, ensuring that the vehicle does not leave the scene. During initialization, a physics engine is set up with a detection volume below the defined bounding volume. Each new point in the pointcloud is automatically registered as a collision object in the physics world. After the octree is populated with an initial set of points (i.e. right after lift-off), the waypoint-generation algorithm is executed as depicted in Fig. \ref{explanation_of_process}:

\begin{enumerate}
     \item Create and initialize sample geometry: spheres of user-defined size are spwaned evenly along the bounding volume's top plane. Increasing the sample geometry's size will create fewer instances, less computational burden (see section \ref{complexity}) and less waypoints, while decreasing the size will find smaller holes the the sampled surface.
     \item Start the physics simulation, allowing the sampling geometry to follow the defined gravity. Whenever a sample geometry instance collides with a point registered in the pointcloud, that event is saved into a special data structure mapping every sample geometry instance to its position in that moment. As this data structure is implemented as a map, later events caused by the same sampling geometry instance replace its previously saved position.
     \item When a sample geometry instance collides with the detection volume, it is removed from the simulation and the position of its last collision on the ground-plane is recorded as a waypoint-marker.
     \item Waypoint-markers are converted to waypoints if there are no other waypoints in close vicinity. When converting markers to waypoints, their height needs to be adjusted, as waypoint-markers tend to spawn on the scenery's ground. The distance $d$, by which markers are lifted before becoming waypoints, is calculated by $d = Range_{lidar} * \frac{SizeSampleGeom_{current}}{SizeSampleGeom_{max}}$. By making the distance from laserscanner to the scanned surface depend on the size of the surface's holes, a natural behavior TODO emerges: The vehicle will first scan larger voids in the surface from further away, then, as the sampling geometry's size is reduced in later iterations of the waypoint generation, come closer to fill smaller defects that couldn't be seen from former sensor poses.
\end{enumerate}

@@

Although any geometry can be used for waypoint generation, using spheres yields three important advantages:
\begin{itemize}
  \item Spheres can be represented only by radius and position, reducing the overall memory requirement (especially when all spheres share a common radius).
  \item Collision detection between points and spheres is a process of low computational complexity, even redundantizing the midphase in collision detection. To detect a collision, it is sufficient to check whether the distance between the sphere's center and the point is smaller than the sphere's radius.
  \item Most importantly, spheres are ideally suited to slip through smaller ``holes'' in the pointcloud, enabling their detection.
\end{itemize}

%Spatially limited by defined exploration space, other input pre-existing sensor information and vehicle position.

%Simulated annealing.

\begin{figure*}[ht]
  \centering
    \includegraphics[width=0.3\textwidth]{images/process_top/1}\hfill
    \includegraphics[width=0.3\textwidth]{images/process_top/2}\hfill
    \includegraphics[width=0.3\textwidth]{images/process_top/3}\vspace{5mm}
    \includegraphics[width=0.3\textwidth]{images/process_top/4}\hfill
    \includegraphics[width=0.3\textwidth]{images/process_top/5}\hfill
    \includegraphics[width=0.3\textwidth]{images/process_top/6}
    \caption{As seen from the top, multiple iterations of waypoint generation lead to continuous scanning at the border between known and unknown environment. Red waypoints are enqueued for scanning, green waypoints have been passed. In the second-to-last figure, sampling geometry of reduced size has allowed finding smaller holes in previously scanned surfaces.}
    \label{figure_process}
\end{figure*}

@@

To assess our algorithms performance, we established the number of points stored in the pointcloud for a given flight-time as the primary metric. As explained in chapter \ref{complexity}, our octree-implementation allows setting a maximum point-density during construction. For the pointcloud storing the surface-reconstruction data, we defined the minimum distance between neighboring points to be $0.1m$. Because the number of points stored correlates directly to the scanned surface-area, measuring the number of points stored over time equals measuring the scanned surface area over time in meaning. All tests were executed in a simulator where the platform's angular velocity remained constant at $120 ^\circ/s$, while its linear velocity was limited to $2.8 m/s$ during all trials. After the scandata was streamed to the basestation using a TCP-connection, it was processed and statistics about flight-time and pointcloud-sizes were logged.

@@

Additionally, the graph's red line depicts our algorithms' behavior: because the pointcloud needs an initial population for waypoint generation, the scanning rate stagnates shortly after launch during the first phase of waypoint generation, then reaches comparable speeds moments later. As

@@

With the algorithm relying on collision detection between large numbers of static and dynamic objects, it is currently limited by the speed of collision detection. Physics-simulation in general and collision detection especially lends itself well to optimization using massively-parallel implementations on GPUs. In the next steps, tests and extensions of the algorithm on GPUs are planned. Here, the questions of whether an increased number of sample geometry instances and their abidance over many iterations of the algorithm improves the waypoint generation is of special interest to the authors.

@@

%\section{ACKNOWLEDGMENTS}
%The authors gratefully acknowledge the contribution of TAMS.

@@

% Octrees
\bibitem{meagher1982}
Donald Meagher, Geometric modeling using octree encoding, {\it Computer Graphics and Image Processing}, 19, 2, pp. 129-147, (1982)

\bibitem{hota2009}
% Dubins Path, no wind
% viele referenzen auf airplanes mit wind uns dubins
Sikha Hota and Debasish Ghose, A Modified Dubins Method for Optimal Path Planning of a Miniature Air Vehicle Converging to a Straight Line Path, American Control Conference, St. Louis, MO, USA, 2009

\bibitem{mcneely2007}
% dubins vehicle und wind, wie kommeich am schnellsten von a nach b.
Rachelle L. McNeely and Ram V. Iyer and Phillip R. Chandler, Tour Planning for an Unmanned Air Vehicle under Wind Conditions, Journal of Guidance, Control, and Dynamics, Vol. 30, No. 5, September-October 2007

\bibitem{joho2007}
Dominik Joho and Cyrill Stachniss and Patrick Pfaff and Wolfram Burgard, Autonomous Exploration for 3{D} Map Learning, {\it {A}utonome {M}obile {S}ysteme {(AMS)}}, Springer, 2007, pp. 22-28

\bibitem{legrand2007}
% macht broadphase per CUDA, gut erklärt in GPU gems 3, chapter 32
Scott Le Grand, Broad-Phase Collision Detection with CUDA, {\it GPU Gems 3, } (2007)

\bibitem{baraff1992}
David Baraff, Dynamic Simulation of Non-Penetrating Rigid Bodies, Cornell University, pp. 52 (1992)

\bibitem{gonzalez_banos2002}
% SLAM does not address the sensor-placement
% Extension to the ArtGallery-Problem
% Sensorkeulen von Kandidatenpunkten berechnen, 2D only
H\'{e}ctor H. Gonz\'{a}lez-Ba\~{n}os and Jean-Claude Latombe, Navigation Strategies for Exploring Indoor Environments, {\it I. J. Robotic Res.} 21(10-11), 829-848 (2002)

\bibitem{haehnel2004}
% Pioneer und Sick zu Outdoor Modellen, 16seiten, kein Wort über next-best-view
Dirk H\"ahnel and Wolfram Burgard and Sebastian Thrun, Learning Compact 3d Models of Indoor and Outdoor Environments with a Mobile Robot, {\it Elsevier Science Special Issue Eurobot} '01, 1-16

\bibitem{strand2009}
% Next-best-view centered. Occupancy grids are most common. cells are not binary, but carry probability.
% grid map mit ground, wall, unexplored
% ein gridcell ist dann guter next best view, wenn von ihm aus ein simulierter scan über möglichst viele groundcells zu unexplored gebiet führt.
% komische metrik, was spielts für ne rolle, wieivel ground der strahl passiert hat.
% das wäre in 3d zu teuer.
Marcus Strand and Ruediger Dillmann, Grid based next best view planning for an autonomous robot in indoor environments

\bibitem{mikrokopterproject}
Mikrokopter project, http://www.mikrokopter.de, 2011

>>>>

\title{\LARGE \bf
On Physics-Based 3D Waypoint Generation for Autonomous Exploration in Aerial Robotics
}

@@

\author{Benjamin Adler and Jianwei Zhang\\
        University of Hamburg\\
        Fachbereich Informatik,\\
        Technische Aspekte Multimodaler Systeme (TAMS)\\
        Hamburg, Germany\\
        {\tt\small adler@informatik.uni-hamburg.de}%
}

@@

This paper presents a novel, physics-based approach to computing waypoints, which are subsequently being used to steer a flying platform in three-dimensional space to map an outdoor environment. Because the algorithm relies on simple physical processes, it is both easy to understand and implement in combination with traditional data structures. Generation of efficient sensor trajectories for maximized information gain operates directly on unorganized point clouds, creating a perfect fit for environment mapping with commonly user LIDAR sensors or time-of-flight cameras. We also present the algorithm by simulating its performance in a virtual outdoor scenario.

@@

Automatic model building has always been one of the most important parts of autonomous robotics. Exploration and mapping - as a subclass of this problem - has been the topic of many papers in recent years. The SLAM problem was first solved for two-dimensional mapping scenarios and later-on expanded to three-dimensional environments.% \cite{shen2011}.

This paper presents a novel approach on increasing the efficiency of such mapping procedures. While the implementations of SLAM have improved greatly in recent years, one aspect of mapping three-dimensional environments was given comparatively little attention: finding the next best view. Typically, SLAM algorithms have been researched and developed on mobile platforms moving on flat surfaces \cite{walthelm2011}, \cite{yamauchi1998}, \cite{joho2007}. This setup does not require a high efficiency in exploration, as it does not inherently impose time constraints. The authors of this paper, however, are implementing exploration and mapping algorithms on a flying platform with a maximum flight time of less than 15 minutes. Aiming to explore as much of the environment as possible within this limited time motivates an efficient way of finding sensor poses (and in extension, sensor trajectories) that enable sensors to deliver as much information as quickly as possible. While repeatedly passing over the unknown environment in scan-line fashion results in very fast scanning, a better approach needs to be found to ensure scanning of surfaces that are shadowed from those sensor trajectories.

This paper is organized as follows: the next section investigates other authors work of recent years in this and related fields. In the subsequent chapters, we first introduce the hardware platform used for testing our implementations and then advance to explaining the basic datastructures our algorithm relies on. After our algorithm is explained and illustrated, we discuss its computational complexity and the results achieved so far. In the final chapter, we share our thoughts on its current limitations and future developments.

@@

Since the introduction of SLAM, much work has been done to refine and extend the algorithm for use in both two- and threedimensional environments. Finding the next best view, though, has never been a part of the SLAM problem \cite{gonzalez_banos2002} and is thus rarely addressed by these works.

Generating safely reachable waypoints from previously acquired sensor data is an extension of next-best-view problem (which itself is an extension to the art-gallery problem \cite{orourke1987}), as it does not include the constraints of safe navigation and instead deals with the generation of isolated viewpoints\cite{chen2008}. As sensor poses in areas between known and unknown environment offer a good compromise between safe reachability and high information gain, frontier-based approaches are a popular method to compute new waypoints in two-dimensional space and are well documented in the literature \cite{gonzalez_banos2002}, \cite{strand2009}, \cite{yamauchi1998}, \cite{shade2011}. Unfortunately, information about exploration boundaries is hard to generate from three-dimensional pointclouds.

As our robot is very maneuverable, the generated waypoints and trajectories can be created with less concern of the platform's ability to catenate them due to motion constraints that other tyes of vehicles (e.g. fixed-wing UAVs) previously entailed \cite{hota2009}. The common problem of wind, manifesting as both opportunity and hindrance, but mostly in the form of further constraints on feasible paths for many UAVs has spawned many contributions in the past \cite{mcgee2005}, \cite{mcneely2007}. Even though wind is still a matter of concern for our platform, it does no longer influence the design of path planning or waypoint generation, but only localization and motion control.

@@

\begin{figure}[thpb]
  \centering
  \includegraphics[width=0.5\textwidth]{images/oktokopter_photo_small}
  \caption{The experimental flying platform with mounted GNSS-antenna/receiver, processor-board, laser scanner and IMU.}
  \label{oktorendering}
\end{figure}

%\cite{mikrokopterproject}
To save time during the construction of our platform, we used an ``Okto 2''-octocopter from the mikrokopter-project as a base for our vehicle (see Fig. \ref{oktorendering}). Its central FlightControl (FC) processor-board is connected to eight brushless-motor controllers (BL) via an $I{^2}C$ bus. Employing its on-board gyros and accelerometers, the FC is programmed to stabilize the platform by itself when no other motion-control-commands are received from either the connected remote-control-receiver or the serial port. The project also developed the NaviControl (NC), an add-on hardware module that sends motion-control-commands (thrust, yaw, pitch and roll each as 8-bit signed integers) to steer the platform to waypoints. We abandoned the idea of using the NC for our purposes, as its hardware adds extra weight and its control-algorithms are neither capable of moving the platform as desired nor open-source. Instead, we fitted an Intel Atom processor-board to the platform. This board has a serial connection to the FC, which is used to send motion-control-commands and receive sensor input.

@@

The laser scanner (Hokuyo UTM 30 LX) rotates at 2400rpm and provides both falling and rising edges after each complete scan of its 270$^\circ$ field of view. This signal is connected to the GNSS receiver, which is configured to emit the GPS Time Of Week (TOW) of the event with microsecond-precision on its USB-connection to the Atom processor-board. With both poses and laser scanner events being timestamped by the GNSS-receiver's time system, we can interpolate the vehicle's pose to the scanner's pose at the time of each scan.

@@

We initially planned to use two laser scanners, one with its scan-plane perpendicular to both the platform's groundplane and its forward-vector for scanning the environment and another scanner oriented to have its scan-plane parallel to the vehicle's ground-plane for collision-avoidance (see Fig. \ref{lidar_setup_pair}). Besides the obvious disadvantages of another 250 grams of payload and $\sim$10W of power consumption, the platform would have had to constantly pitch during flight to ensure reliable obstacle scanning and detection. Superimposing the pitching motion has proven to result in considerable amounts of work for the motors on the platform's front and back, reducing flight time. The non-constant angular pitching speed in this setup also necessitates a more precise interpolation of the vehicle's to the scanner's pose.

To avoid these drawbacks, we instead mounted a single laser scanner as depicted in Fig. \ref{lidar_setup_single}. By permanently yawing the platform in-flight, a single scanner captures both geometry for surface reconstruction and collision avoidance, provided that the world is static and the platform does not move further than the laser scanners' range within one rotation. Given a constant angular yawing-speed, the disadvantage of precise interpolation of the platform's pose becomes less problematic.
% TODO: speed < lidar-range * drehgeschwindigkeit ?

@@

%\cite{meagher1982}
In order to capture sensor data representing arbitrarily complex geometry and detail, we decided to implement our algorithm on an octree-based data-structure. Octrees are well suited for storing information in non-uniform resolution and the process of storing and retrieving data can easily be executed in parallel, allowing for optimization of parts of our algorithms. Besides its position, each point in the octree also stores its brightness (i.e. intensity as measured by the laser scanner) and a vector pointing from that point back to the sensor. As the platform's measured orientation is less precise that its measured position, this vector's length will be used lateron to assign points recorded from further distance (and thus suffering a greater impact from platform-orientation errors) a lesser weight in the following surface reconstruction.

@@

\begin{figure*}[htb]
  \centering
  \subfigure[]{
    \label{method_1}
    \includegraphics[width=0.43\textwidth]{images/process_side/1}
  }\hfill
  \subfigure[]{
    \label{method_2}
    \includegraphics[width=0.43\textwidth]{images/process_side/2}
  }
  \subfigure[]{
    \label{method_3}
    \includegraphics[width=0.43\textwidth]{images/process_side/3}
  }\hfill
  \subfigure[]{
    \label{method_4}
    \includegraphics[width=0.43\textwidth]{images/process_side/4}
  }
  \caption{Waypoint generation, orthogonal projection; a) depicts the initial setup: the bounding volume is blue, detection volume is red, vehicle position is black, pointcloud for geometry reconstruction is grey and pointcloud for waypoint generation is red. b) shows sampling geometry interacting with pointcloud. In c), geometry hitting the red volume below is converted to a waypoint. The last figure d) shows generated waypoints in red for an initial pointcloud population. Please see http://tinyurl.com/physwpt for an animation of the process.}
  \label{explanation_of_process}
\end{figure*}

Three-dimensional environment mapping is often implemented using laser scanners and time-of-flight cameras, so unorganized point clouds are a very common type of sensor data. Any sensor generating spatial occupancy information can be used with our algorithm, as the pointcloud is its only sensory input. Finding the next best view in such data can be very hard, as it does not supply any information about geometric structures such as corners, edges, surfaces and normals. In contrast to many other algorithms (\cite{joho2007}, \cite{walthelm2011}), ours does not require such information.

Prior to the mapping process, the user creates a bounding volume as a representation of the region-of-interest around the vehicle's starting position, thereby defining the environment that is to be mapped. All generated viewpoints will be constrained to this volume and its vicinity, ensuring that the vehicle does not leave the scene. During initialization, a physics engine is set up with a detection volume below the defined bounding volume. Each new point in the pointcloud is automatically registered as a collision object in the physics world. After the octree is populated with an initial set of points (i.e. right after lift-off), the waypoint generation algorithm is executed as depicted in Fig. \ref{explanation_of_process}:

\begin{enumerate}
     \item Create and initialize sample geometry: spheres of user-defined size are spwaned evenly along the bounding volume's top plane. Increasing the sample geometry's size will create fewer instances, less computational burden (see section \ref{complexity}) and less waypoints, while decreasing the size will find smaller holes the the sampled surface.
     \item Start the physics simulation, allowing the sampling geometry to follow the defined gravity. Whenever a sample geometry instance collides with a point registered in the pointcloud, that event is saved into a special data structure mapping every sample geometry instance to its position in that moment. As this data structure is implemented as a map, later events caused by the same sampling geometry instance replace its previously saved position.
     \item When a sample geometry instance collides with the detection volume, it is removed from the simulation and the position of its last collision on the ground-plane is recorded as a waypoint marker.
     \item Waypoint markers are converted to waypoints if there are no other waypoints in close vicinity. When converting markers to waypoints, their height needs to be adjusted, as waypoint markers tend to spawn on the scenery's ground. The distance $d$, by which markers are lifted before becoming waypoints, is calculated by $d = Range_{lidar} * \frac{SizeSampleGeom_{current}}{SizeSampleGeom_{max}}$. By making the distance from laser scanner to the scanned surface depend on the size of the surface's holes, a natural behavior emerges: The vehicle will first scan larger voids in the surface from further away, then, as the sampling geometry's size is reduced in later iterations of the waypoint generation, come closer to fill smaller defects that couldn't be seen from former sensor poses.
\end{enumerate}

@@

% \cite{sun2002}, 
Depending on size, number and shape of sampling geometry, this method simulates exposing the scanned surface to a fluid pushing through the model. The idea of generating watertight models from previously generated datasets is rather common and already well researched \cite{hornung2006}, \cite{sormann2007}, so the novelties here are:

\begin{itemize}
 \item testing the currently collected data for watertightness in-flight and using the result as a termination criterion for exploration.
 \item testing for watertightness using a physical simulation and its real-time implementation directly on sensor-data without additional pre-processing.
 \item compromise between exploration time and quality of resulting model can be changed quickly and in-flight by changing sampling geometry size.
 \item the algorithm's gracefully degraded behavior when size and number of sampling geometry cannot be scaled to achieve simulation with near-fluid behavior due to constrained computational resources.
\end{itemize}

Although any geometry can be used for waypoint generation, using spheres yields three important advantages:

\begin{itemize}
  \item Spheres can be represented only by radius and position, reducing the overall memory requirement (especially when all spheres share a common radius).
  \item Collision detection between points and spheres is a process of low computational complexity, even redundantizing the midphase in collision detection. To detect a collision, it is sufficient to check whether the distance between the sphere's center and the point is smaller than the sphere's radius.
  \item Most importantly, spheres are ideally suited to slip through smaller ``holes'' in the pointcloud, enabling their detection.
\end{itemize}

\begin{figure*}[htb]
  \centering
  \subfigure[]{
    \label{process_1}
    \includegraphics[width=0.3\textwidth]{images/process_top/1}
  }
  \subfigure[]{
    \label{process_2}
    \includegraphics[width=0.3\textwidth]{images/process_top/2}
  }
  \subfigure[]{
    \label{process_3}
    \includegraphics[width=0.3\textwidth]{images/process_top/3}
  }
  \subfigure[]{
    \label{process_4}
    \includegraphics[width=0.3\textwidth]{images/process_top/4}
  }
  \subfigure[]{
    \label{process_5}
    \includegraphics[width=0.3\textwidth]{images/process_top/5}
  }
  \subfigure[]{
    \label{process_6}
    \includegraphics[width=0.3\textwidth]{images/process_top/6}
  }
  \caption{As seen from the top, multiple iterations of waypoint generation lead to continuous scanning at the border between known and unknown environment. Red waypoints are enqueued for scanning, green waypoints have been passed. In figures a) to d), sample spheres with $r=3m$ have been used to quickly explore the landscape, while in figure e), sampling geometry of reduced size ($r=1.5m$) has allowed finding smaller holes in previously scanned surfaces.}
  \label{figure_process}
\end{figure*}

@@

To assess our algorithms performance, we established the number of points stored in the pointcloud for a given flight-time as the primary metric. As explained in chapter \ref{complexity}, our octree-implementation allows setting a maximum point-density during construction. For the pointcloud storing the surface-reconstruction data, we defined the minimum distance between neighboring points to be $0.1m$. Because the number of points stored correlates directly to the scanned surface-area, measuring the number of points stored over time equals measuring the scanned surface area over time in meaning. All tests were executed in a simulator where the platform's angular velocity remained constant at $120 ^\circ/s$, while its linear velocity was limited to $2.8 m/s$ during all trials. After the scandata was streamed to the basestation using a TCP-connection, it was processed and statistics about flight-time and pointcloud sizes were logged.

@@

Additionally, the graph's red line depicts our algorithms' behavior: because the pointcloud needs an initial population for waypoint generation, the scanning rate stagnates shortly after launch during the first phase of waypoint generation, then reaches comparable speeds as the vehicle starts passing waypoints on the border between known and unknown environment. In this phase of flight, our algorithm is as efficient as a manually planned collision free optimal path, with the obvious advantage of having been generated automatically. Compared to exploring in scanline-fashion, the true strength in our method lies in the fact that on-line re-planning occurs such that when no more waypoints can be generated, gradually smaller sampling geometry is used to improve on ever smaller deficiencies in the reconstructed surface. This yields a truly dynamic and efficient scanning procedure that adapts well to many different kinds of outdoor scenarios.

@@

With the algorithm relying on collision detection between large numbers of static and dynamic objects, it is currently limited by the speed of collision detection. Physics-simulation in general and collision detection especially lend themselves well to optimization using massively-parallel implementations on GPUs. In the next steps, tests and extensions of the algorithm on GPUs are planned. Here, the questions of whether an increased number of sample geometry instances (up to a method that completely fills the user-defined scanning region with sample geometry) and their abidance over many iterations of the algorithm improves the waypoint generation is of special interest to the authors.

@@

% TODO shortened
%Finally, we plan to move our research's focus from waypoint generation to the problems of generating safely navigatable paths between them, dynamically replanning paths when necessary and discarding those that cannot be reached safely.

@@

% Octrees
%\bibitem{meagher1982}
%Donald Meagher, Geometric modeling using octree encoding, {\it Computer Graphics and Image Processing}, 19, 2, pp. 129-147, (1982)

\bibitem{hota2009}
% 1mal
% Dubins Path, no wind
% viele referenzen auf airplanes mit wind uns dubins
Sikha Hota and Debasish Ghose, A Modified Dubins Method for Optimal Path Planning of a Miniature Air Vehicle Converging to a Straight Line Path, American Control Conference, St. Louis, MO, USA, 2009

\bibitem{mcneely2007}
% 1mal
% dubins vehicle und wind, wie kommeich am schnellsten von a nach b.
Rachelle L. McNeely and Ram V. Iyer and Phillip R. Chandler, Tour Planning for an Unmanned Air Vehicle under Wind Conditions, Journal of Guidance, Control, and Dynamics, Vol. 30, No. 5, September-October 2007

% 1 mal
\bibitem{chen2008}
Shengyong Chen and Y. F. Li and Jianwei Zhang and Wanliang Wang, Active Sensor Planning for Multiview Vision Tasks, Springer Verlag Berlin Heidelberg, 2008

\bibitem{mcgee2005}
% 1mal
Timothy G. McGee and Stephen Spry and J. Karl Hedrick, Optimal path planning in a constant wind with a bounded turning rate, Proc. of the AIAA Guidance, Navigation and Control Conference and Exhibit, San Francisco, CA, August 2005.

\bibitem{joho2007}
% 2mal
Dominik Joho and Cyrill Stachniss and Patrick Pfaff and Wolfram Burgard, Autonomous Exploration for 3{D} Map Learning, {\it {A}utonome {M}obile {S}ysteme {(AMS)}}, Springer, 2007, pp. 22-28

\bibitem{legrand2007}
% 2 mal
% macht broadphase per CUDA, gut erklärt in GPU gems 3, chapter 32
Scott Le Grand, Broad-Phase Collision Detection with CUDA, {\it GPU Gems 3, } (2007)

\bibitem{baraff1992}
% 1mal
David Baraff, Dynamic Simulation of Non-Penetrating Rigid Bodies, Cornell University, pp. 52 (1992)

\bibitem{gonzalez_banos2002}
% 1mal
% SLAM does not address the sensor-placement
% Extension to the ArtGallery-Problem
% Sensorkeulen von Kandidatenpunkten berechnen, 2D only
H\'{e}ctor H. Gonz\'{a}lez-Ba\~{n}os and Jean-Claude Latombe, Navigation Strategies for Exploring Indoor Environments, {\it I. J. Robotic Res.} 21(10-11), 829-848 (2002)

%\bibitem{haehnel2004}
% Pioneer und Sick zu Outdoor Modellen, 16seiten, kein Wort über next-best-view
%Dirk H\"ahnel and Wolfram Burgard and Sebastian Thrun, Learning Compact 3d Models of Indoor and Outdoor Environments with a Mobile Robot, {\it Elsevier Science Special Issue Eurobot} '01, 1-16

\bibitem{strand2009}
% 1mal
% Next-best-view centered. Occupancy grids are most common. cells are not binary, but carry probability.
% grid map mit ground, wall, unexplored
% ein gridcell ist dann guter next best view, wenn von ihm aus ein simulierter scan über möglichst viele groundcells zu unexplored gebiet führt.
% komische metrik, was spielts für ne rolle, wieivel ground der strahl passiert hat.
% das wäre in 3d zu teuer.
Marcus Strand and Ruediger Dillmann, Grid based next best view planning for an autonomous robot in indoor environments

%\bibitem{mikrokopterproject}
%Mikrokopter project, http://www.mikrokopter.de, 2011
